---
title: "Data Visualisation with ggplot2 - Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
My focus in this analysis is to provide strategy recommendations that will lead to revenue growth. Some of the question I intent to answer include:

* How many bookings were canceled?
* What was the booking ratio between resort hotels and city hotels?
* What was the percentage of booking for each year?
* Which was the busiest month for hotels?
* Most guests came from which country?
* How long do most people stay in hotels?
* Which was the most booked accommodation type (Single, Couple, Family)?

### b) Defining the Metric for Success
The solutions to the following questions will help us answer my research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context

I will perform data exploration, cleaning and analysis on the provided dataset to develop strategies that will help the Hilton International Hotels management decide on
the strategies that will grow the revenue.

### d) Recording the Experimental Design

* Data reading
* Data exploration
* Data cleaning
* Data preparation
* Analysis

### e) Data Relevance

The provided dataset has enough features that will help answer the defined research questions.


## 2. Reading the Data

```{r}
# Install and load TidyVerse
library(tidyverse)
# YOUR CODE GOES BELOW
install.packages("reshape2")

# loading the package
library(reshape2)
# Load the data below
# --- 
dataset_url = "https://raw.githubusercontent.com/wambasisamuel/DE_Week03_Tuesday/main/hotel_bookings.csv"
# --- 
# YOUR CODE GOES BELOW
bookings_df <- read.csv(dataset_url)
```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
head(bookings_df, 5)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
tail(bookings_df, 5)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
sample_n(bookings_df, 10, replace = TRUE) 
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
dim(bookings_df)  
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
str(bookings_df)
```

```{r}

```

Record your general observations below:

* The dataset has 119390 records and 32 columns(features) 
* The columns contain either numeric or string data


## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.
[Link](https://www.sciencedirect.com/science/article/pii/S2352340918315191)

## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
colSums(is.na(bookings_df))

```

We observe the following from our dataset:

The dataset four missing values for the children column
Observation 2

```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
bookings_df[duplicated(bookings_df), ]
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
colSums(is.na(bookings_df))
```

We observe the following from our dataset:

There are no null values in all columns

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW
is.null((bookings_df)) 
```

We observe the following from our dataset:

There are no rows having all values null
Observation 2

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
library(tidyverse)
# YOUR CODE GOES BELOW
install.packages("reshape2")

# loading the package
library(reshape2)

# Select the desired columns
heatmap_df <- bookings_df %>%
  select(is_canceled, stays_in_weekend_nights, stays_in_week_nights, is_repeated_guest)

# Create a correlation matrix
corr_matrix <- cor(heatmap_df, method="s")
head(corr_matrix)

# Create a table with long form
corr_df <- melt(corr_matrix)
head(corr_df)

# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) + 
  geom_tile(color = "black") + 
  geom_text(
    aes(label = round(value, 2)), 
    color = "white"
  ) +
  coord_fixed() + 
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_gradient2(
    low = "blue", 
    high = "red",
    mid = "white", 
    limit = c(-1,1)
  ) + 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

```

We observe the following from our dataset:

Observation 1
Observation 2

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 
bookings_df <- select(bookings_df, -c("company"))
```

From the data variable description we see that the Distribution Channel categoricy that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#
bookings_df$agent[is.na(bookings_df$agent)] <- "TO"
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
bookings_df <- bookings_df[!(bookings_df$adults == 0 | bookings_df$babies == 0 | bookings_df$children == 0),]
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
rounded_mean <- round(mean(bookings_df$children, na.rm = TRUE))
rounded_mean
bookings_df$children[is.na(bookings_df$children)] <- rounded_mean
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
colSums(is.na(bookings_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW

# We dropped company column earlier
bookings_df <- bookings_df %>%
  transform(children = as.integer(children), agent = as.integer(agent))

#is.na(bookings_df$children)
#str(bookings_df)
#is.na(bookings_df$agent)
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot

# Summary table
cancelled_df <- bookings_df %>%
  filter(is_canceled == 1) %>%
  count(hotel, is_canceled, sort = TRUE, name = "Count")

# Plotting the chart
ggplot(cancelled_df, aes(x = hotel, y = Count)) +
  geom_col(
    stat="identity", width=0.5, fill = "#b8ff33"
  )  +
  labs(
    title = "Cancelled Bookings per Hotel", 
    x = "Hotels", 
    y = "Cancellations"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) 
```


```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel
# Summary table
ratio_df <- bookings_df %>%
  count(hotel, is_canceled, sort = TRUE, name = "Count")

# Plotting the chart
ggplot(ratio_df, aes(x = hotel, y = Count)) +
  geom_col(
    stat="identity", width=0.5, fill = "#ffd433"
  )  +
  labs(
    title = "Booking Ratio between Resort and City Hotel", 
    x = "Hotels", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Comic Sans MS"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) 
```


```{r}
# 3. What was the percentage of booking for each year?
# ---
# Summary table
piechart_df <- bookings_df %>%
  group_by(arrival_date_year) %>%
  summarise(bookings=n())

piechart_df$arrival_date_year <- as.character(piechart_df$arrival_date_year)


# Plotting the chart
ggplot(data=piechart_df, aes(x="", y=bookings, fill=arrival_date_year)) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") + 
  geom_text(
    aes(label = paste0(round(bookings / sum(bookings) * 100, 1), "%")), position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set2") +   
  labs(
    fill = "Year", 
    title = "Bookings per Year"
  ) +
  theme_void()
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# Summary table
months_df <- bookings_df %>%
  group_by(arrival_date_month) %>%
  summarise(bookings=n()) %>%
  arrange(desc(bookings)) 

ggplot(months_df, aes(x = arrival_date_month, y = bookings)) +
  geom_col(
    stat="identity", width=0.5, fill = "#33ffd1"
  )  +
  labs(
    title = "Bookings per Month", 
    x = "Month", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Mono"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The top three busiest months are August, July and December


```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
top_countries_df <- bookings_df %>%
  group_by(country) %>%
  summarise(bookings=n()) %>%
  arrange(desc(bookings)) 

ggplot(top_countries_df, aes(x = country, y = bookings)) +
  geom_col(
    stat="identity", width=0.5, fill = "#33ffd1"
  )  +
  labs(
    title = "Bookings per Country", 
    x = "Country", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Mono"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The top 3 countries are PRT, ESP and FRA

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
total_stay_df <- bookings_df %>%
  filter(is_canceled == 0) %>%
  mutate(total_nights = rowSums(across(c(stays_in_weekend_nights,stays_in_week_nights)))) %>%
  group_by(total_nights) %>%
  summarise(counts=n()) %>%
  arrange(desc(counts))

total_stay_df$total_nights <- as.character(total_stay_df$total_nights)

# Plotting the chart
ggplot(total_stay_df, aes(x = total_nights, y = counts)) +
  geom_col(
    stat="identity", width=0.5, fill = "#33ffd1"
  )  +
  labs(
    title = "Number of Stays", 
    x = "Number of nights", 
    y = "Counts"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Mono"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) 
```

Most people stay for one, two, or three nights


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---

singles_df <- bookings_df %>%
  filter(is_canceled == 0, adults == 1, children == 0, babies ==0) %>%
  group_by(country) %>%
  summarise(bookings=n()) %>%
  arrange(desc(bookings))

couples_df <- bookings_df %>%
  filter(is_canceled == 0, adults == 2, children == 0, babies ==0) %>%
  group_by(country) %>%
  summarise(bookings=n()) %>%
  arrange(desc(bookings))

families_df <- bookings_df %>%
  filter(is_canceled == 0, adults == 1, children == 0, babies ==0) %>%
  group_by(country) %>%
  summarise(bookings=n()) %>%
  arrange(desc(bookings))
```

### 5.b) Recommendations
From the above analysis, below are our recommendations:



## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# ---
# YOUR GOES BELOW
rates_df <- bookings_df %>%
  select(arrival_date_month, adr) %>%
  group_by(arrival_date_month) %>%
  summarize(avg_rate = mean(adr, na.rm = TRUE)) %>%
  arrange(desc(avg_rate))

ggplot(rates_df, aes(x = arrival_date_month, y = avg_rate)) +
  geom_col(
    stat="identity", width=0.5, fill = "#33ffd1"
  )  +
  labs(
    title = "Monthly Average Daily Rate", 
    x = "Month", 
    y = "Avg  Rate"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Our observations:

Hotels should provide offers for months where the average daily rate is high. This is in the months of March, July, August and September. This can increase bookings for these months

How does this observation tie to our solution?

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?

The dataset provided fits the analysis

b). Do we need other data to answer our question?

No other data is needed to answer the questions

c). Did we have the right question?

The question asked was right

